---
title: "Heli-attack times"
author: "Michael Storey, MichaelJones, Mitchell Stephen, Michael May"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    code_folding: hide
---

```         
To Do:
  * Create waterpoints/reservoir/lakes layer
  * Classify water points (long line, medium heavy etc.)
  * Amend script point creation to exclude demand points in the ocean/NSW
  * Improve leaflet maps (scalable symbols? Or IDW raster or something?)
  * Display test plots 3 across
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#set local library path so R knows where to look. You will also need to run this line below before downloading new packages, just due to some cenitex issues.
#.libPaths("D:/R/R-Packages")

# load necessary packages
library(sf)
library(dplyr)
library(leaflet)
library(leaflet.extras)
library(htmltools)

```

### Setup

-   Define all inputs and variables

Remainder of script should run without requiring modification

```{r Chunk 1: Setup}

# path locations
output_path <- "C:/Projects/20230829_HeliWaterPoints/waterpoints/outputs"
input_path <- "C:/Projects/20230829_HeliWaterPoints/waterpoints/data/"

# shapefile names
dem_file <- "DEM_SE_Tas_mosaic.tif"
water_poly_file <- "reservoir_all_new.shp"
new_dams_file <- "potential_dam_locations_14052020.shp"

# specify which water points to include (not required, but fix if needed)
# water_poly_filter <- (NAME == ... | ... | ...)

# helicopter speeds in kilometers per hour
all_empty_kmh <- 148
full_ascend_kmh <- 74
full_level_kmh <- 148 #same as full descending

# point grid size for analysis in metres
point_grid_size<-5000

# path sample density (metres between samples from reservoirs to points)
line_sample_point_dist <- 100

# choose whether to include new dams (1) or not (0). Don't use 1 under a 1k grid, will take too long.
include_new_dams <- 0

# construct full file paths
dem <- file.path(input_path, dem_file)
water_poly_full_path <- file.path(input_path, water_poly_file)
new_dams_full_path <- file.path(input_path, new_dams_file)

```

### Create demand grid

-   Load shapefiles
-   Transform coordinate reference system (CRS),and define the variable.
-   Create a point grid across region (ie encompassing the reservoirs of interest).
-   Plot grid points black and water in blue

```{r Chunk 2: Create demand grid, echo=TRUE, message=FALSE, warning=FALSE}

# load and project shapefiles
water_poly<-st_read(water_poly_full_path,quiet=TRUE) %>% st_transform(3111,quiet=TRUE) 
  # %>% filter(water_poly_filter)

new_dams<-st_read(new_dams_full_path,quiet=TRUE)  
st_crs(new_dams)<-3111
point_grid<- water_poly %>% st_make_grid(point_grid_size,what='centers') %>% st_as_sf()
water_points<-st_cast(water_poly,"POINT")

# add new dams to water points if include_new_dams variable set to 1
if(include_new_dams == 1){water_points<-st_union(water_points,new_dams) %>% st_cast("POINT")}

```

### Create traversal lines

-   Draw a line from every point in grid to its closest water source.

```{r Chunk 3: Create traversal lines, message=FALSE, warning=FALSE}

# create lines from every grid point to nearest water point
nn<-nngeo::st_nn(point_grid,water_points,k=1,progress = FALSE)
near_lines<-nngeo::st_connect(point_grid,water_points,k=1,progress = FALSE)
near_lines<-st_as_sf(near_lines) 
near_lines$line_id<-seq.int(nrow(near_lines))

water_points$wp_id<-seq.int(nrow(water_points))
point_grid<-st_join(point_grid,water_points,join=st_nearest_feature)
near_lines$wp_id<-point_grid$wp_id


```

### Embed DEM into traversal lines

-   For every line from point grid to water, add a vertex every 100 meters and turn vertices into points (transferring line id to new point feature class)
-   Sample the DEM for every line vertex point, and select the point with the maximum elevation.
-   Sample the DEM for elevation of start point (where fire would be) and end point (water point)
-   Calculate the difference in elevation between start, end, and max elevation

```{r Chunk 3: Embed DEM, message=FALSE, warning=FALSE}

# densify lines (make vertices every <line_sample_point_dist> metres)
near_lines_dense<- near_lines%>% smoothr::densify(max_distance=line_sample_point_dist)
near_lines_vertices<-st_cast(near_lines_dense,"POINT",do_split = TRUE,warn = FALSE)

# load dem
raster_dem<-raster::raster(dem)

# extract water point elevation
water_points$dem_wp<-raster::extract(raster_dem,st_transform(water_points,4326))

# extract elevation of traversal vertices
point_grid$dem_start<-raster::extract(raster_dem,st_transform(point_grid,4326))
near_lines$dem_start<-point_grid$dem_start

#near_lines_vertices$dem_path <- raster::extract(raster_dem, st_transform(near_lines_vertices, 4326, quiet=TRUE))
#max_dem_vertex <- near_lines_vertices %>% group_by(line_id) %>% filter(dem_path==max(dem_path)) %>% mutate(n=n()) %>% mutate(positionInCategory = 1:n()) %>% filter(positionInCategory == 1)

#near_lines$peak_dem <- max_dem_vertex$dem_path

# join height of water point to lines
d<-data.frame(water_points %>% select(wp_id,dem_wp)) %>% select(-geometry)
near_lines<- near_lines %>% left_join(d,by="wp_id") 

# calculate height differences
#near_lines$st_peak_diff<-near_lines$peak_dem- near_lines$dem_start
#near_lines$wp_peak_diff<-near_lines$peak_dem- near_lines$dem_wp

# densify the near_lines
near_lines_dense <- near_lines %>%
  smoothr::densify(max_distance = line_sample_point_dist)

# extract vertices from densified lines and add elevation data
near_lines_vertices <- st_cast(near_lines_dense, "POINT", do_split = TRUE, warn = FALSE)
near_lines_vertices$dem_path <- raster::extract(raster_dem, st_transform(near_lines_vertices, 4326, quiet = TRUE))


# create a function to check if a vertex doesn't have a higher elevation anywhere before and after it
check_elevation <- function(line_id, elevations) {
  n <- length(elevations)
  condition_met <- rep(TRUE, n)  # Initialize all points as meeting the condition

  if (any(is.na(elevations))) {
    cat("NA values detected in line_id:", line_id, "elevations:", elevations, "\n")
    return(condition_met)  # If any elevation is NA, return as is
  }

  for (i in 1:(n-1)) {
    max_before <- max(elevations[1:(i-1)])
    max_after <- max(elevations[(i+1):n])
    # check if current vertex is lower than any point on both sides
    if ((max_before > elevations[i]) && (max_after > elevations[i])) ||
      #cat("Fail - Line: ", line_id, "Point: ", i, elevations[i], "lower than: ", max_before, "and ", max_after, "\n")
      condition_met[i] <- FALSE  # Mark the specific point as not meeting the condition
    }
    else {
      cat("Pass - Line: ", unique(line_id), "Point: ", i, elevations[i], 
          "higher than: ", max_before, "or ", max_after, "\n")
    }
  }
  return(condition_met)
}

# function to remove inneficient parts of flight paths (run AFTER check elevation, not before)
check_efficiency <- function (line_id, elevations) {
  n <- length(elevations)
  condition_met <_ rep(TRUE, n)
  
  for (p in 1:(n-2)) {
    current_point <-
    for (n in )
  }
}

# Filter vertices based on elevation condition and set elevation for the remaining vertices
filtered_near_lines_vertices <- near_lines_vertices %>%
  group_by(line_id) %>%
  mutate(elevation_condition_met = check_elevation(line_id, dem_path)) %>%
  ungroup() %>%
  filter(elevation_condition_met) %>%
  select(-elevation_condition_met)


# Extract the necessary columns from filtered_near_lines_vertices
point_data <- filtered_near_lines_vertices %>%
  mutate(x = st_coordinates(.)[, 1], y = st_coordinates(.)[, 2]) %>%
  select(line_id, dp_z = dem_start, wp_z = dem_wp, x, y, z = dem_path)

#calculate_travel_time and distance

# New dataframe to store segment data
segment_data <- data.frame(
  line_id=integer(),
  dist_m=double(),
  elev_change_m=double(),
  time_s=integer(),
  stringsAsFactors=FALSE)

# Initialize variables to store results
line_ids <- unique(point_data$line_id)
distances <- numeric(length(line_ids))
travel_times <- numeric(length(line_ids))

for (i in 1:length(line_ids)) {
  
  line_id <- line_ids[i]
  cat("Calculating Line", line_id, "\n")
  
  # Subset data for the current line_id
  line_data <- point_data[point_data$line_id == line_id, ]
  
  for (p in 2:length(line_data)) {
    diffsq_x <- (line_data[p]$x - line_data[p-1]$x)^2
    diffsq_y <- (line_data[p]$y - line_data[p-1]$y)^2
    diffsq_z <- (line_data[p]$z - line_data[p-1]$z)^2
    
    dist_m <- sqrt(sum(diffsq_x, diffsq_y, diffsq_z))
    elev_change_m <- (line_data[p]$y - line_data[p-1]$y)
    time_s<-1000
    line_id_data <- array(line_id, dist_m, elev_change_m, time_s)
    line_id_data
    
    segment_data <- rbind(segment_data, line_id_data)
  }
}
    
    
    ###### UP TO HERE!!!!
    # Calculate inclination/declination adjustment
  }
  
  # Calculate distance between consecutive points
  distances[i] <- sum(st_distance(line_data$geometry[-1], line_data$geometry[-nrow(line_data)]))
  
  # Calculate travel time based on elevation changes
  elevation_diff <- diff(line_data$z)
  travel_time <- sum(elevation_diff / full_level_kmh)  # Assuming constant speed
  travel_times[i] <- travel_time
}

# Create a data frame with results
result_df <- data.frame(line_id = line_ids, distance = distances, travel_time = travel_times)


# Call the function with your point_data table
travel_time_results <- calculate_travel_time(point_data)

# 5. Create new lines between remaining vertices
#new_lines <- filtered_near_lines_vertices %>%
#  group_by(line_id) %>%
#  summarise(
#    geom = st_cast(st_union(x), "LINESTRING")
#  ) %>%
#  ungroup()

# calculate distance for each line segment using Pythagoras
new_lines$length <- st_length(new_lines$geom)
```

### Draw test plots

-   Plot 1 illustrates the demand grid and water point locations. Grid points are black circles and water is blue
-   Plot 2 adds traversal lines from demand grid points to nearest water point (brown lines)
-   Plot 3 includes the highest point vertex in red

```{r Chunk 4: Draw test plots, message=FALSE, warning=FALSE}

# draw test plot of water point locations
plot(st_geometry(point_grid))
plot(st_geometry(water_points),col='blue',add=TRUE)

# draw test plot with lines between demand points and nearest water point
plot(st_geometry(point_grid))
plot(st_geometry(water_points),col='blue',add=TRUE)
plot(st_geometry(near_lines),col="brown",add=TRUE)

# draw test plot including max DEM point in red
plot(st_geometry(point_grid))
plot(st_geometry(water_points),col='blue',add=TRUE)
plot(st_geometry(near_lines),add=TRUE,col='brown')
plot(st_geometry(max_dem_vertex),add=TRUE,col = 'red')

```

### Calculate helicopter travel times

-   Calculate the distances between start to high dem point, and high dem point to end (water)
-   Using distances, height differences and speeds (all_empty_kmh\<-148,full_ascend_kmh\<-74,full_level_kmh\<-148) to calculate time on different segments of line. At the moment this assumes heli ascends and descends vertically (not on an angle)

```{r Chunk 5: Helicopter travel times, message=FALSE, warning=FALSE}

# get distance between high point and start and end points
high_point_to_wp<-st_distance(max_dem_vertex ,water_points)
near_lines$peak_to_wp<-as.numeric(apply(high_point_to_wp, 1, FUN=min))
near_lines$dist<-as.numeric(st_length(near_lines)) 
near_lines$peak_to_start<-near_lines$dist - near_lines$peak_to_wp

# calculate slope
near_lines$slp_wp_dist<-sqrt((near_lines$wp_peak_diff^2)+(near_lines$peak_to_wp^2))
near_lines$slp_start_dist<-sqrt((near_lines$st_peak_diff^2)+(near_lines$peak_to_start^2))

# calculate times
near_lines <- near_lines %>% mutate(time_full_ascend = (wp_peak_diff/1000)/full_ascend_kmh) %>% 
  mutate(time_full_level = (dist/1000)/full_level_kmh) %>% 
  mutate(time_full_descend = (st_peak_diff/1000)/full_level_kmh) %>% 
  mutate(time_empty_ascend = (st_peak_diff/1000)/all_empty_kmh) %>% 
  mutate(time_empty_level = (dist/1000)/all_empty_kmh) %>% 
  mutate(time_empty_descend = (wp_peak_diff/1000)/all_empty_kmh) %>% 
  mutate(total_time = round((time_full_ascend + time_full_level +time_full_descend +
           time_empty_ascend + time_empty_level + time_empty_descend)*60,1) + 2) #plus 2 minutes for fill time

point_grid$total_time<-near_lines$total_time

# calculate times using pythagoras (ie assuming heli flies on angles over)
near_lines <- near_lines %>% mutate(time_full_ascend_slope = (slp_wp_dist/1000)/full_ascend_kmh) %>% 
  mutate(time_full_descend_slope = (slp_start_dist/1000)/full_level_kmh) %>% 
  mutate(time_empty_ascend_slope = (slp_start_dist/1000)/all_empty_kmh) %>% 
  mutate(time_empty_descend_slope = (slp_wp_dist/1000)/all_empty_kmh) %>% 
  mutate(total_time_slope = round((time_full_ascend_slope +time_full_descend_slope +
           time_empty_ascend_slope + time_empty_descend_slope)*60,1) + 2) #plus 2 minutes for fill time

point_grid$total_time<-near_lines$total_time
point_grid$total_time_slope<-near_lines$total_time_slope

#st_write(point_grid, paste0(output_path, "point_grid.shp"), delete_layer = TRUE)
#st_write(near_lines, paste0(output_path, "near_lines.shp"), delete_layer = TRUE)
#st_write(max_dem_vertice, paste0("D:/TEMP/", "max_dem.shp"), delete_layer = TRUE)
#MJ writes point_grid and near_lines layer at the end

```

### Helicopter travel time map (simple)

-   Interactive map of time to closest reservoir (in minutes). Click on a point to see the minutes
-   This map assumes helicopters ascend and descend vertically, and fly horizontally in-between

```{r Chunk 6: Map 1, message=FALSE, warning=FALSE}

#distances
point_grid_wgs<-st_transform(point_grid,4326) %>% filter(total_time > 0) %>% as("Spatial")
qpal<- colorNumeric(palette = "YlOrBr",domain = seq(50))

map<-leaflet(point_grid_wgs) %>%
  addTiles() %>% 
  addCircleMarkers(color = ~qpal(total_time), popup=~htmlEscape(total_time), radius=10, fillOpacity = 1) %>% addPolygons(data=as(st_transform(water_poly,4326),"Spatial"))%>% 
  addLegend("bottomright", pal = qpal, values = ~total_time,title = "time in minutes",opacity = 1)# %>%
  #addCircleMarkers(data=as(st_transform(new_dams,4326),"Spatial"),color = "blue", radius=10, fillOpacity = 1) 
map 

```

### Helicopter travel time map (alternative)

-   This map assumes helicopters fly on angles over peaks

```{r message=FALSE, warning=FALSE}

#distances
point_grid_wgs<-st_transform(point_grid,4326) %>% filter(total_time_slope > 0) %>% as("Spatial")
qpal<- colorNumeric(palette = "YlOrBr",domain = seq(50))

map<-leaflet(point_grid_wgs) %>%
  addTiles() %>% addCircleMarkers(color = ~qpal(total_time_slope), popup=~htmlEscape(total_time_slope), radius=5, fillOpacity = 0.5) %>% addPolygons(data=as(st_transform(water_poly,4326),"Spatial"))%>% 
  addLegend("bottomright", pal = qpal, values = ~total_time_slope,title = "time in minutes",opacity = 1)# %>%
  #addCircleMarkers(data=as(st_transform(new_dams,4326),"Spatial"),color = "blue", radius=10, fillOpacity = 1) 
map 

```

### Export final data to ArcGIS

-   To do

```{r message=FALSE, warning=FALSE}

# To do
st_write(point_grid, paste0(output_path, "point_grid.shp"), delete_layer = TRUE)
st_write(near_lines, paste0(output_path, "near_lines.shp"), delete_layer = TRUE)

```
